---
title: 'Survival Analysis: Ovarian Cancer'
author: "Nesma Magdi"
date: "October 18, 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Greg)
library(dplyr)
library(tidyr)
library(finalfit)
library(ggplot2)
library(moments)
library(e1071)
library(dummies)
library(dplyr)
library(tidyr)
library(finalfit)
library(dummies)
library(moments)
library(corrplot)
library(scatterplot3d)
library(ggplot2)
library(ggplot2)
library(caret)
library(GGally)
library(ggthemes)
library(broom)
library(dplyr)
library(bindrcpp)
library(caTools)
library(rattle)
library(RColorBrewer)
library(nnet)
library(rpart.plot)
library(dichromat)
library(Rmpfr)
library(digest)
library(convertr)
library(microbenchmark)
library(varhandle)
library(hablar)
library(magrittr)
library(survminer)
library(markovchain)
library(survival)
library(expss)
library(AMR)
library(ggpubr)
library(gplots)
library(devtools)
library(dmetar)
library(grDevices)
library(Hmisc)
library(rattle)
library(reshape)
library(magrittr)
library(ggpubr)
library(gridExtra)
library(PerformanceAnalytics)
library(highcharter)
library(car)
```

## Ovarian Cancer:

Survival in a randomised trial comparing two treatments for ovarian cancer.
Format

Treatment of patients with advanced ovarian carcinoma (stages IIIB and IV) using either cyclophosphamide alone (1 g/m2) or cyclophosphamide (500 mg/m2) plus adriamycin (40 mg/m2) by iv injection every 3 weeks each produced partial regression in approximately one third of the patients.

```{r ovarian, message=FALSE, warning=FALSE}

data("ovarian")
attach(ovarian)

Status <- ifelse(fustat == 0,"Death", "Censord")
Progression <- ifelse(resid.ds == 1, "No Progression", "Progression")
Treatment <- ifelse(rx == 1, "Endoxan alone", "Endoxan + Adria")
Performance <- ifelse(ecog.ps == 1, "High", "Low")
dummy_Age <- ifelse(age >= 50 ,"Old","Young")

ex <- data.frame(Status, Progression, Treatment, Performance, dummy_Age)

Ovarian <- cbind(ovarian, ex)

Ovarian$dummy_Age <- factor(Ovarian$dummy_Age)
Ovarian$Status<- factor(Ovarian$Status)
Ovarian$Progression <- factor(Ovarian$Progression)
Ovarian$Treatment <- factor(Ovarian$Treatment)
Ovarian$Performance <- factor(Ovarian$Performance)


write.csv(Ovarian, file="ovarian.csv")
ovarian <- read.csv(choose.files(), header = T)
attach(ovarian)

```

## Descrition:

```{r , echo=FALSE}

Abbreviation <- c("futime", "fustat", "age", "resid.ds", "rx", "ecog.ps")

Description <- c("Survival or Censoring time", "Censoring status", "Age in years", "Residual disease present", "Treatment Group", "ECOG Performance Status")

Data <- as.data.frame(Description , Abbreviation)
Data
```

```{r , echo=TRUE}

summary(ovarian)

```

## First, Analysis of Variance:

#1- Age:

```{r warning=FALSE, , echo=TRUE}

AGE_mean <- as.data.frame(ovarian %>%
              summarise(count = n(), mean = mean(age, na.rm = T), sd = sd(age, na.rm = T)))
AGE_mean

summary(ovarian$age)
ggplot(ovarian, aes(x= age)) + 
  geom_histogram(aes(y=..count..), color = "black", fill="firebrick4")

hist (age , col="bisque" , freq=FALSE)
meanx <- mean( age )
sdx <- sd (age)
curve (dnorm(x , mean=mean(x) , sd=sd(x) ) ,
from=min(age ) , to=max(age) , add=TRUE,
lty=1, col="Black " , lwd=3)
kernel <- density (age)
points ( kernel$x  , kernel$y , type="l " ,
lty=2, col="blue " , lwd=3)
legend ("topleft" , c ("Normal " , "Kernel") , lty=c (1 ,2) ,
bty="n" , cex=.9 , col=c (" black " , " blue "))



ggboxplot(age, xlab = "Age") +
  labs(x = "Age", y = "Count", main = "Age Normality", fill = "age") +
  theme_economist()

skewness(age)
kurtosis(age)

shapiro.test(age)
t.test(age, mu=56.17,  conf.level = 0.95)

```

  Mean age of population of Intrest is 56.17 and standard Deviation is 10.1.
  Age of Patients: From 38.89 To 74.5 years.
  
  The Distribution is moderatly Skewed & approximatly Symmetric, and Kurtosis is in acceptable range for bing normally distributed.

From the output, the p-values= 0.4259 > the significance level 0.05, implying that the distribution of the data are not significantly different from the normal distribution.

In other words, we can assume the normality, Normally Distributed.

 The p-value of one sample t.test is 0.9982, which is greater than the significance level alpha = 0.05. We can conclude that average Ages is not significantly different from there mean with a p-value = 0.9982.
 
===============================================================

#2- Progression:

```{r warning=FALSE, , echo=TRUE}

Count_Prog <- table(ovarian$Progression)
Percent_Prog <- prop.table(Count_Prog)*100

Prog_persent <- as.data.frame(cbind(Count_Prog, Percent_Prog))
Prog_persent


ggplot(ovarian,aes(Progression, fill= Progression)) +
  geom_bar(stat="count") +
  theme_economist()

chisq.test(Count_Prog)
```

The p-value of the Chi.Square test is 0.4328, which is greater than the significance level alpha = 0.05. We can conclude that Prgoression is significantly commonly distributed with a p-value = 0.4328.

===============================================================

#3- ECOG Performance Status:

```{r warning=FALSE, , echo=TRUE}

Count_per <- table(ovarian$Performance)
Percent_per <- prop.table(Count_per)*100

per_persent <- as.data.frame(cbind(Count_per, Percent_per))
per_persent

ggplot(ovarian,aes(Performance, fill= Performance)) +
  geom_bar(stat="count") +
  theme_economist()

chisq.test(Count_per)
```

The p-value of the Chi.Square test is 0.6949, which is greater than the significance level alpha = 0.05. We can conclude that ECOG Performance Status is significantly commonly distributed with a p-value = 0.6949.

********************************************************************************************************************************************************************************************************************************************************

## Second, Survival Analysis:

# Kaplan-Meier Estimator:

```{r warning=FALSE, , echo=TRUE}

survobj <- with(ovarian, Surv(futime, fustat))

surviv <- survfit(survobj ~ 1, data = ovarian)
summary(surviv)

plot(surviv, xlab = "Survival Time in Days", ylab = "% Surviving", yscale = 100, main = "Survival Distribution (Overall)")

ggsurvplot(surviv, data = ovarian, pval = TRUE)

```
 
## 1- Compare the Survival Distribution Between Old and Young Female Patients (Survival Curves):

```{r warning=FALSE, , echo=TRUE}

surviv1 <- survfit(survobj ~ dummy_Age, data = ovarian)
surviv1

plot(surviv1, xlab = "Survival Time in Months", ylab = "% Surviving", yscale = 100, 
     col = c("dodgerblue4", "firebrick4"),
     main = "Survival Distribution Between Different ages")
     
     legend("topright", title = "Ages", c("Old", "Young"),
     fill= c("dodgerblue4", "firebrick4"))
     
ggsurvplot(surviv1, data = ovarian, pval = TRUE)

survdiff(survobj ~ dummy_Age, data = ovarian)
```

p-value = 0.1, This means Median Survival Time of old and young patients is not Significantly different.

===============================================================
 
## 2- Compare the Survival Distribution of Different Progression Status (Survival Curves):

```{r warning=FALSE, , echo=TRUE}

surviv2 <- survfit(survobj ~ ovarian$Progression, data = ovarian)
surviv2

plot(surviv2, xlab = "Survival Time in Months", ylab = "% Surviving", yscale = 100, 
     col = c("dodgerblue4", "firebrick4"),
     main = "Survival Distribution Between Different Progression")
     legend("topright", title = "Progression", c("Yes", "No"),
     fill= c("dodgerblue4", "firebrick4"))
     
ggsurvplot(surviv2, data = ovarian, pval = TRUE)

survdiff(survobj ~ ovarian$Progression, data = ovarian)
```

p-value = 0.6, This means Median Survival Time between different progression status is not Significantly different.

===============================================================
 
## 3- Compare the Survival Distribution of Different Performance Status (Survival Curves):

```{r warning=FALSE, , echo=TRUE}

surviv3 <- survfit(survobj ~ ovarian$Performance, data = ovarian)
surviv3

plot(surviv3, xlab = "Survival Time in Months", ylab = "% Surviving", yscale = 100, 
     col = c("dodgerblue4", "firebrick4"),
     main = "Survival Distribution Between Different Performance")
     legend("topright", title = "Performance", c("Low", "High"),
     fill= c("dodgerblue4", "firebrick4"))
     
ggsurvplot(surviv3, data = ovarian, pval = TRUE)

survdiff(survobj ~ ovarian$Performance, data = ovarian)
```

p-value = 0.5, This means Median Survival Time between different performance status is not Significantly different.

===============================================================
 
## 4- Compare the Survival Distribution of Different Treatments (Survival Curves):

```{r warning=FALSE, , echo=TRUE}

surviv4 <- survfit(survobj ~ ovarian$Treatment, data = ovarian)
surviv4
summary(surviv4)

plot(surviv4, xlab = "Survival Time in Months", ylab = "% Surviving", yscale = 100, 
     col = c("dodgerblue4", "firebrick4"),
     main = "Survival Distribution Between Treatments")
     legend("topright", title = "Performance", c("Endoxan + Adria", "Endoxan alone"),
     fill= c("dodgerblue4", "firebrick4"))
     
ggsurvplot(surviv4, data = ovarian, pval = TRUE)

survdiff(survobj ~ ovarian$Treatment, data = ovarian)
```

p-value = 0.3, This means Median Survival Time after using  different Treatment Regimens is not Significantly different.
Though, patients receiving Cyclophosphamide alone are doing better in the first month of follow-up. 

===============================================================
 
## 4- Progression Free Survival:

```{r warning=FALSE, , echo=TRUE}

stat1 <- survfit(Surv(futime, fustat) ~ 1, data = ovarian)
Prog1 <- survfit(Surv(futime, resid.ds) ~ 1, data = ovarian)

Fitting <- list(OS = stat1, PFS = Prog1)
Fitting
ggsurvplot_combine(Fitting, ovarian)
ggsurvplot(Fitting, data = ovarian, pval = TRUE)
```

********************************************************************************************************************************************************************************************************************************************************

## Third, Cox Regression:

# refrence = Endoxan alone:

```{r warning=FALSE, , echo=TRUE}

survobj <- with(ovarian, Surv(futime, fustat))

Model1 <- coxph(survobj ~ rx, data = ovarian)
summary(Model1)

```
 
1- Regression Coeffecient = -0.5964, Negative sign means Risk of Death is lower when using Cyclophophamide alone.
2- Patients using Cyclophophamide alone has lower risk of death than using Cyclophophamide & Adriamycin.
3- The effect size of age as a covarience = 0.55, This mean:
Taking Cyclophophamide alone reduce hazard of death by 45%.

```{r warning=FALSE, , echo=TRUE}

progobj <- with(ovarian, Surv(futime, resid.ds))

Model11 <- coxph(progobj ~ rx, data = ovarian)
summary(Model11)

```
 
1- Regression Coeffecient = -0.5104, Negative sign means Risk of Progression is lower when using Cyclophophamide alone.
2- Patients using Cyclophophamide alone has lower risk of Progression than using Cyclophophamide & Adriamycin.
3- The effect size of age as a covarience = 0.6, This mean:
Taking Cyclophophamide alone reduce hazard of Progression by 40%.

===============================================================

# Survival per Treatment from other variables:

* Hazard of Death from ovarian cancer considering th Covarience:

```{r warning=FALSE, , echo=TRUE}

Model2 <- coxph(survobj ~ rx+  age + resid.ds + ecog.ps, data = ovarian)
summary(Model2)
ggforest(Model2, data = ovarian)

```

1- Older Patients has higher risk of death, this risk is significant (p-value = 0.00777)
2- likelihood ratio test Significance decreased from 0.3 to 0.002.


# Evaluate Proportional Hazard Assumption:

```{r warning=FALSE, , echo=TRUE}

cox.zph(Model2)
par(mfrow = c(2, 4))
plot(cox.zph(Model2), data = ovarian)
```

# By using the Schoenfeld residuals against the transformed time:

From the output above, the test is not statistically significant for each of the covariates, and the global test
is also not statistically significant. Therefore, we can assume the proportional hazards.

1- rho: Pearson's correlation between the scaled Shoenfeld residuals and Time. The value of ecog.ps has
higher corralation (0.4845) from over covariate, and resid.ds has the least corrlation (-0.1417). All in positive direction except age and resid.ds in Negative direction.

2- chisq: The correlation coefficient follows a chi-square distribution and the statistic is present in the second column.

3- p-value: P value is given for each covariate.

4- For the global test there is no correlation and NA is entered into the cell.

************************************************************************************************************************************************************************************************************************************************************

# Progression Free Survival per Treatment from other variables:

* Hazard of Progression from ovarian cancer considering th Covarience:

```{r warning=FALSE, , echo=TRUE}

progobj <- with(ovarian, Surv(futime, resid.ds))

Model22 <- coxph(progobj ~ rx+  age + fustat + ecog.ps, data = ovarian)

summary(Model22)

ggforest(Model22, data = ovarian)

```

1- Older Patients has higher risk of Progression, this risk is significant (p-value = 0.00403)
2- Different in Status has higher risk of Progression, this risk is significant (p-value = 0.00303)
3- High Performance Status has Lower risk of Progression, this risk is significant (p-value = 0.02030)

# Evaluate Proportional Hazard Assumption:

```{r warning=FALSE, , echo=TRUE}

cox.zph(Model22)
par(mfrow = c(2, 4))
plot(cox.zph(Model22), data = ovarian)
```

# By using the Schoenfeld residuals against the transformed time:

From the output above, the test is not statistically significant for each of the covariates, and the global test
is also not statistically significant. Therefore, we can assume the proportional hazards.

1- rho: Pearson's correlation between the scaled Shoenfeld residuals and Time. The value of ecog.ps has higher corralation (0.194) from over covariate, and age has the least corrlation (-0.300). All in positive direction except age and fustat in Negative direction.

2- chisq: The correlation coefficient follows a chi-square distribution and the statistic is present in the second column.

3- p-value: P value is given for each covariate.

4- For the global test there is no correlation and NA is entered into the cell.

************************************************************************************************************************************************************************************************************************************************************



